A note on observation processes in epidemic models

### 2020-02-13 ms/response.tex

\revtext{Authors did some work in this direction, but I believe more work is needed to address their question. In particular, authors will need to use a model that incorporates realistic distributions of latent and infectious periods. They can accomplish this using the linear chain tricks or alternative integro-differential equation models. Then they can generate simulated data from that "true/realistic" model and go ahead and estimate $\mathcal R_0$ using simple models as they do here.}

\revtext{Having said this, I would like to point out that I am sure there are a number of papers out there that have look at similar questions in the context of different infectious diseases, etc. It'd be good to make sure the relevant literature is considered and appropriately cited.}

The reviewer may have misconstrued our goals in this paper. We were \emph{not} exploring the effects of the process delay distributions (i.e. latent and infectious periods), and their mis-specification, on epidemiological dynamics and parameter estimation. Rather, we were exploring the effects of reporting delays (and their mis-specification) on parameter estimation.

However, reporting delays do depend implicitly on the process delays. We have added the following paragraph in the Discussion with appropriate citations to address this point:

``Distributions of reporting time delays implicitly depend on the latent and infectious 
period distributions. For example, when cases are reported
at the end of infectious periods,
the delay distribution is equivalent to the convolution of the latent and
infectious period distributions. These distributions also play important roles in 
shaping the epidemiological dynamics,
including the probability and size of an outbreak, the stability and persistence of a system,
and the  occurrence of dynamical transitions.
Assumptions about latent and infectious period distributions can also have large effects on the estimates of the basic reproductive number.''

@@

\revtext{I have one comment that the authors may wish to consider, although this doesn't rise to the level of a critical flaw that must be corrected.  It seemed to me that the Discussion lost sight of the fact that statistical models often have to make simplifying assumptions.  Despite these simulation results, it seems to me that making a simplifying assumption about the process generating the incidence data may not be a terrible thing to do with real data.  When it comes to fitting statistical models to noisy data, models that are "more realistic", in the sense of incorporating more of the underlying mechanism, do not necessarily lead to improved inference.  One has to choose their battles with respect to what one tries to capture in a model, and what one simplifies away.  With real data, it is entirely unclear whether model fit would be improved by fitting a model with a more complicated description of the process generating the incidence data, and it may very well be that such a
question can only be answered on a case-by-case basis.  In the absence of such understanding, it seems to me that the Discussion could strike a somewhat more measured tone regarding the merits of simpler vs. more complicated models.}

We agree. We did not intend to say that modelers should always strive to build the more complex models. We have revised the final paragraph of the Discussion to better reflect our intentions:

``All statistical models rely on simplifying assumptions.
Given a range of possible assumptions, modelers have to
decide which aspects of a system to capture 
at the cost of simplifying other aspects.
However, some assumptions are often taken for granted.
Epidemic modelers tend to focus on assumptions about the
transmission process; processes through which cases are reported are often 
overlooked and simplified.
Our study shows that a seemingly negligible change in the assumptions of an epidemic
model can affect the inference of 
infectious disease transmission.
Although the amount of bias introduced from the misspecification of the observation
model can be small (e.g., less than 5\% in our examples), modelers
trying to make serious predictions may want to control for potential biases.
We caution disease modelers to be more careful about their modeling decisions as they may have unexpected consequences on the inference.''
